Estimating duration in the presence of misclassification

### 2024-03-26 cis-imperfect-testing.tex
proof ch 6 inc new figures

In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
However, the initial application of this framework produced an implausibly short estimate.
In this chapter I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.

@@

Applying the method of \cref{E-perf-test} to CIS data estimates fewer episodes with a duration of less than three weeks, compared to \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having frequent data collection here (see \cref{E-biology-data:sec:ataccc,E-ATACCC:sec:discussion}).
This suggests an issue in the estimation using CIS data.
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/CIS_perfect}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

Comparing repeated data simulations, using the set-up of \cref{E-perf-test:sec:simulation-study}, to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
Specifically, if $j$ is a single positive episode then $j$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
Additionally, short episodes are very likely to be undetected, which means that the design of the CIS cannot rule out many of them occurring.
% This produced the same pattern, that is estimating too many short episodes (not shown).

@@

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of \numprint{1000}, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data are shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this section, I introduce a probabilistic model for the generation of false negatives.
The simplest model for false negatives is a constant, non-zero probability of returning a negative test when detectable.
That is the test sensitivity, $\psens$, is less than 1.

Under this model, each test result in a detectable individual is an iid Bernoulli random variable with a probability $\psens$ of producing a positive.
As previously, tests in individuals without ongoing infection episodes are always negative.
I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.

@@

The CIS data provides evidence for a test sensitivity declining over the course of an episode.
\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
\begin{enumerate}
    \item All positive results are true positives.
    \item All intermittent negatives are false negatives.
    \item The negative at the end of an episode is a possible false negative.
    \item All other negatives are true negatives.
\end{enumerate}
Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie the first positive test in the episode).
This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
However, the general trend is clear.
\begin{figure}
  \includegraphics{cis-imperfect-testing/test-sens-bound}
  \caption[Bounding test sensitivity using CIS data]{
    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
  }
  \label{imperf-test:fig:bounding-cis-sensitivity}
\end{figure}

To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
  \label{imperf-test:eq:variable-test-sensitivity}
\end{equation}
where $t$ is the number of days since infection.
Therefore, this model consists of a linear decline in test sensitivity before a constant minimum point.

I simulate data under this model by, for each test in a detectable individual, calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
The result of the test is then a Bernoulli random variable, independent of all other considerations.
Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.

@@

Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data will involve summing over the possible episode start times.

@@

For tractability, assume that the negative test bounding the start of the episode, on day $l_j^{(b)}-1$, is a true negative.
This assumption is reasonable because, since a positive test follows at $r_j^{(b)}$, the negative at $l_j^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
True negatives occur with probability 1, and hence this test does not contribute to the likelihood.

Denote the set of the remaining test times to consider as $\sched'_{i(j)}$.
$\sched'_{i(j)}$ is the set of times individual $i(j)$ was tested between $l_j^{(b)}$ and $r_j^{(e)} + 1$ inclusive.
Denote their results by $\vec{y}_j = \{ y_{i(j)}(t) \ssep t \in \sched'_{i(j)} \}$ where $y_{i(j)}(t) = 1$ if the test on day $t$ in individual $i(j)$ is positive and 0 otherwise.

As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_j, l^{(e)}_j]$, because this period starts and ends with a positive test.
This includes all $t \in \sched'_{i(j)}$ except $t = r_j^{(e)}+1$.
Therefore, the test results $\vec{y}_j$, except the test at $r_j^{(e)}+1$, are either true positives or false negatives.

@@

First, if $E_j \leq r_j^{(e)}$, meaning that the test at $r_j^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
% In this case, the test at $r_j^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&\prob(\vec{y_j}, E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{y_j}, l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
&= \prob(\vec{y_j} \mid l_j^{(e)} \leq E_j \leq r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

Second, if $E_j > r_j^{(e)}$.
In this case, the test at $r_j^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
To avoid having to consider tests after $r_j^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_j^{(e)}$.
Taking the same approach as before:
\begin{align}
&\prob(\vec{y_j}, E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{y_j} \mid E_j > r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b_j + 2)
\label{imperf-test:eq:ll-ei-gt-ri}
\end{align}

These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_j}$.
First, augment the data with $b_j$, and split into the cases just discussed:
\begin{align}
p_{ia}'
=& \prob(\vec{y_j} \mid p_\text{sens}, \vec{\theta}) \\
=& \sum_{b = l_j^{(b)}}^{r_j^{(b)}} \left( \prob(\vec{y_j}, E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_j}, E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
  &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
=& \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \\ & \ \times \sum_{b = l_j^{(b)}}^{r_j^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right) \\ 
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
=& \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \\
  & \times \sum_{b_j = l_j^{(b)}}^{r_j^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_j^{(e)} - b_j + 2) \right) \\
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.

If $\psens$ is fixed (\ie has a point prior) and $p(b_j \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
\begin{align}
p_{ia}'
&\propto \sum_{b = l_j^{(b)}}^{r_j^{(b)}} S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_j^{(e)} - b + 2).
\label{imperf-test:eq:pia-prime-constant}
\end{align}

@@

Therefore, this mechanism causes episode $j$ to be undetected if all the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:sec:prob-undetected}. That is $b_j > \min(\sched_{i(j)})$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
    \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j]$.
      Note that the lower bound here is exactly the bound on $e_j$ in the previous condition.
      Equivalently, $\tau_{\sched_{i(j)}}(b_j) + 1 \leq d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
    \item A false negative occurs on day $\tau_{\sched_{i(j)}}(b_j) + b_j$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_j = b_j > \min(\sched_{i(j)})$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_{i(j)}}(b_j) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1
    \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b_j) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b_j) + 1) \right) (1 - \psens).
\end{align}
Summing over $b_j$ for $\min \sched_{i(j)} < b_j \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{perf-test:eq:piu}, gives:
\begin{align}
\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right).
\end{align}

@@

I simulate datasets under four conditions: a constant $\psens$ of 0.6, 0.8, or 1.0 (the latter being the same as the simulated data in \cref{E-perf-test:sec:simulation-study}), or the varying test sensitivity model proposed at the end of \cref{imperf-test:sec:simulate}.
The simulation is otherwise unchanged from that in \cref{E-perf-test:sec:simulation-study}.
The number of single positive episodes in each of these conditions, compared to the number observed in the CIS data, is shown in \cref{imperf-test:fig:sim-single-pos}.

For each simulated dataset, I estimate the survival function using the likelihood proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise, I refer to it as misspecified.
However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
The amount that the simplifying assumptions are violated increases as $\psens$ decreases.

I use either the independent (vague) or model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}).
These were shown to be the best performing priors in \cref{E-perf-test:sec:results}.
Therefore, there are a total of 24 possible scenarios, although not all are shown.
These are each combination of: 4 simulated datasets, with varying $\psens$; 3 values of $\psens$ in inference; and 2 priors for the hazard.

>>>>

In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
However, the initial application of this framework produced an implausibly short estimate.
In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.

@@

Applying the method of \cref{E-perf-test} to CIS data estimates a median duration of less than 14 days, several days shorter than the estimates from \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to ATACCC's frequent data collection (see \cref{E-biology-data:sec:ataccc,E-ATACCC:sec:discussion}).
This suggests an issue in the estimation using CIS data.
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/CIS_perfect}
  \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}. \label{imperf-test:fig:problem-cis-estimates}}
\end{figure}

To investigate this further, I simulated \numprint{1000} datasets using the setup of \cref{E-perf-test:sec:simulation-study}.
The simulations produced far fewer single positive episodes than seen in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause shorter estimates because the lower bound on the length of time the episode is one day, the shortest possible.
Specifically, if $j$ is a single positive episode then $j$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
Additionally, short episodes are very likely to be undetected, which means that the design of the CIS cannot rule out many of them occurring.
% This produced the same pattern, that is estimating too many short episodes (not shown).

@@

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives.
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of \numprint{1000}, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data are shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this section, I introduce a probabilistic model for the generation of false negatives.
Recall, from \cref{E-biology-data:sec:measurement-error}, that the test sensitivity, $\psens$, is the probability of a positive test result given that the individual is detectable.
The simplest model for false negatives is a constant $\psens < 1$.

Under this model, each test in a detectable individual generates an iid $\Ber(\psens)$ result where a one indicates a positive and zero indicates a negative.
As previously, tests in undetectable individuals are always negative.
I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.

@@

The CIS data provides evidence for the test sensitivity declining over the course of an episode.
\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
\begin{enumerate}
    \item All positive results are true positives.
    \item All intermittent negatives are false negatives.
    \item The negative at the end of an episode is a possible false negative.
    \item All other negatives are true negatives.
\end{enumerate}
The classification assumes that the probability of two false negative tests at the end of an episode is negligible.
As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
This can be bounded by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
I calculate the test sensitivity for each day, where the day of the first positive result in an episode is day zero.
Day zero is an unknown number of days after the episode's latent true beginning time.
% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie the first positive test in the episode).
I generate broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
However, the general trend is clear.
\begin{figure}
  \includegraphics{cis-imperfect-testing/test-sens-bound}
  \caption[Bounding test sensitivity using CIS data]{
    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
  }
  \label{imperf-test:fig:bounding-cis-sensitivity}
\end{figure}

The following model allows simulating with a varying test sensitivity, allowing assessment of the inference procedure's sensitivity to this.
Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
  \label{imperf-test:eq:variable-test-sensitivity}
\end{equation}
where $t$ is the number of days since infection.
The model consists of a linear decline in test sensitivity before a constant minimum point.

I simulate data under this model by, for each test in a detectable individual, calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
The result of the test is then a Bernoulli random variable, independent of all other considerations.
Incorporating this test sensitivity into the simulation gives much more similar results to the CIS data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
Furthermore, it does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.

@@

Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data involves summing over the possible episode start times.

@@

For tractability, assume that the negative test bounding the start of the episode, on day $l_j^{(b)}-1$, is a true negative.
This assumption is reasonable because, since a positive test follows at $r_j^{(b)}$, the negative at $l_j^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
True negatives occur with probability 1, and hence this test does not contribute to the likelihood.

Denote the set of the remaining test times to consider as $\sched'_{i(j)}$, the set of times individual $i(j)$ was tested between $l_j^{(b)}$ and $r_j^{(e)} + 1$ inclusive.
Denote their results by $\vec{y}_j = \{ y_{i(j)}(t) \ssep t \in \sched'_{i(j)} \}$ where $y_{i(j)}(t) = 1$ if the test on day $t$ in individual $i(j)$ is positive and 0 otherwise.

As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_j, l^{(e)}_j]$, because this period starts and ends with a positive test.
This includes all $t \in \sched'_{i(j)}$ except $t = r_j^{(e)}+1$.
Therefore, the test results $\vec{y}_j$, except the test at $r_j^{(e)}+1$, are either true positives or false negatives.
This gives $t_+ = \sum_{t \in \sched'_{i(j)}} y_{i(j)}(t)$ true positives and $f_- = \sum_{t \in \sched'_{i(j)}} (1 - y_{i(j)}(t))$ false negatives.

@@

First, the case when $E_j \leq r_j^{(e)}$.
In this case, the test at $r_j^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
% In this case, the test at $r_j^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&\prob(\vec{y_j}, E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{y_j}, l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
&= \prob(\vec{y_j} \mid l_j^{(e)} \leq E_j \leq r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

Second, the case when $E_j > r_j^{(e)}$.
In this case, the test at $r_j^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
To avoid having to consider tests after $r_j^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_j^{(e)}$.
Taking the same approach as before:
\begin{align}
&\prob(\vec{y_j}, E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{y_j} \mid E_j > r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b_j + 2)
\label{imperf-test:eq:ll-ei-gt-ri}
\end{align}

These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the probability of observing $\vec{y_j}$ given that an infection occurs in individual $i$.
First, augment the data with $b_j$, and split into the cases just discussed:
\begin{align}
p_{ia}'
=& \prob(\vec{y_j} \mid p_\text{sens}, \vec{\theta}) \\
=& \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( \prob(\vec{y_j}, E_j \leq r_{j(i)}^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_j}, E_j > r_{j(i)}^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
  &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
 & \times \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \\ 
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
  & \times \sum_{b_j = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_j + 2) \right) \\
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$ (see \cref{perf-test:eq:pia}).

If $\psens$ is fixed (\ie has a point prior) and $p(b_j \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
\begin{align}
p_{ia}'
&\propto \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2).
\label{imperf-test:eq:pia-prime-constant}
\end{align}

@@

Therefore, this mechanism causes episode $j$ to be undetected if all the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:sec:prob-undetected}. That is $\min(\sched_{i(j)}) < b_j \leq \max(\sched_{i(j)})$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
    \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j]$.
      Note that the lower bound here is exactly the bound on $e_j$ in the previous condition.
      Equivalently, $\tau_{\sched_{i(j)}}(b_j) + 1 \leq d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
    \item A false negative occurs on day $\tau_{\sched_{i(j)}}(b_j) + b_j$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_j = b_j > \min(\sched_{i(j)})$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_{i(j)}}(b_j) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1, b_j \leq \max(\sched_{i(j)}),
    \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b_j) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b_j) + 2) \right) (1 - \psens).
\end{align}
Summing over $b_j$ for $\min \sched_{i(j)} < b_j \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{perf-test:eq:piu}, gives:
\begin{align}
\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right).
\end{align}

@@

Denote by $\psenss$ the test sensitivity value used when simulating the dataset, with $\psens = v$ indicating the varying test sensitivity was used.
Denote by $\psensi$ the constant test sensitivity value assumed during inference.

I use one dataset from \cref{imperf-test:sec:simulate} for each $\psenss \in \{ 0.6, 0.8, 1.0, v \}$ ($\psenss = 1$ is the simulated data in \cref{E-perf-test:sec:simulation-study}).
The simulation is otherwise unchanged from that in \cref{E-perf-test:sec:simulation-study}.
The number of single positive episodes in each of these conditions, compared to the number observed in the CIS data, is shown in \cref{imperf-test:fig:sim-single-pos}.

For each simulated dataset, I estimate the survival function using the posterior density proposed in \cref{imperf-test:sec:modelling} and assuming each $\psensi \in \{0.6, 0.8, 1.0\}$.
If $\psenss = \psensi$, I refer to $\psens$ as being correctly specified; otherwise, I refer to it as misspecified.
However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
The amount that the simplifying assumptions are violated increases as $\psenss$ decreases.

I use either the independent (vague) or model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}).
These were shown to be the best performing priors in \cref{E-perf-test:sec:results}.
Therefore, there are a total of 24 possible scenarios, although not all are shown.
These are each combination of: 4 simulated datasets, with the different values of $\psenss$; 3 values of $\psensi$; and 2 priors for the hazard.

When $\psens = 0.8$ and is correctly specified, the model recovers the true survival time well (see \cref{imperf-test:fig:constant-test-sensitivity}(B)).
The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplifying assumptions, moving the estimated survival function closer to its true survival time.
However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}(A)).
This is likely caused by too large a violation of the simplifying assumptions made in \cref{imperf-test:sec:modelling}.
\begin{figure}
  % \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}}
  \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}
  \caption[Simulation study results with constant test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with $\psens$ correctly specified.
    The true survival time is shown in black.
  }
  \label{imperf-test:fig:constant-test-sensitivity}
\end{figure}
